From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/discussion.tex
Finalise

Coherence theorems are famous problems in (higher) category theory, and Mac Lane's coherence
theorem~\cite{maclaneNaturalAssociativityCommutativity1963,joyalBraidedTensorCategories1993,gurskiInfiniteLoopSpaces2013}
for monoidal categories is a particular one. The use of rewriting theory and proof assistants to establish coherence
theorems for various higher categorical structures has a long history as well,
see~\cite*{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
$(A \eqv B) \eqv (A \id_{\UU} B)$.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case).

There are other, different approaches to computing with univalence,
in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
Cubical Type
Theory\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.

@@

%%%%%%%%%%%%%%
\paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe structures such as groups
or rings. A specific group or ring is a model of the appropriate algebraic theory. Algebraic theories are usually
\emph{presented} in terms of logical syntax, that is, as first-order theories whose signatures allow only functional
symbols, and whose axioms are universally quantified equations. In his seminal thesis,
\citet{lawvereFUNCTORIALSEMANTICSALGEBRAIC1963} defined a presentation-free categorical notion of universal algebraic
structure, called a Lawvere theory. Programming Languages, such as the $\lambda$-calculus, can be viewed as algebraic
structures with variable-binding operators, which can be formalised using second-order algebraic
theories~\cite{fioreSecondOrderAlgebraicTheories2010}, or algebraic theories with closed
structure~\cite{hylandClassicalLambdaCalculus2017}, called $\lambda$-theories, making the $\lambda$-calculus the
presentation of the initial $\lambda$-theory $\Lambda$. Our family of reversible languages $\PiLang$ have been presented
as first-order algebraic
2-theories~\cite{cohenCoherenceRewriting2theories2009,bekeCategorificationTermRewriting2011,yanofskySyntaxCoherence2000},
which are a categorification of algebraic theories. The types $\zerot$ and $\onet$ are nullary function symbols, the
type formers $+$ and $\times$ are binary function symbols, the 1-combinators are invertible reduction rules, and the
2-combinators are equations or coherence diagrams of compositions of reduction rules. Just like models of Lawvere
theories are given by algebras of (finitary) monads on $\SetCat$, models of 2-theories are given by algebras of 2-monads
on $\CatCat$. The particular one related to our development is the free symmetric monoidal completion 2-monad.

@@

Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories
were studied by \citet{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a
groupoid and the proof of its universal property has been considered
by~\citet{piceghelloCoherenceMonoidalGroupoids2020}. Free commutative monoids in type theory have been studied by
\citet{gylterudMultisetsTypeTheory2020}, and using HoTT by \citet{choudhuryFinitemultisetConstructionHoTT2019}. The free
symmetric monoidal groupoid $\FSM[A]$ over a groupoid $A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be
presented as an algebraic 2-theory using 1-HITs. These HITs and the proof of their universal property have been
considered by \citet{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}. The proof of
the universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal
categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The
existence of the proof is folklore. We have produced a proof and formalised it in constructive type theory.

A stronger universal property that $\BFin$ is biinitial in the 2-category of rig categories, was established
in~\cite{elguetaGroupoidFiniteSets2021}.

@@

%%%%%%%%%%%%%%
\paragraph{Permutations} One might argue that finding formal systems for expressing various flavors of computable
functions has been a major focus of logic and computer science since its inception. Permutations, being at the core of
reversible computing, are an interesting class of functions, the bijective ones, for which there is no sound and
complete formal system. Our work develops such a formal system bringing in all the associated benefits of
syntactic calculi, most notably, their calculational flavor. Thus instead of comparing two reversible programs by
extensional equality of the underlying bijective functions, a calculus offers more nuanced techniques that can enforce
additional intensional constraints on the desired equality relation.

>>>>

Coherence theorems are famous problems in category theory, and Mac Lane's coherence
theorem~\cite{maclaneNaturalAssociativityCommutativity1963,joyalBraidedTensorCategories1993,gurskiInfiniteLoopSpaces2013}
for monoidal categories is a particular one. The use of rewriting and proof assistants to prove coherence theorems
for higher categorical structures has a long history
see~\cite*{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
$(A \eqv B) \eqv (A \id_{\UU} B)$.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,
different approaches to computing with univalence,
in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
Cubical Type
Theory\cite*{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.

@@

%%%%%%%%%%%%%%
\paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe structures such as groups
or rings. A specific group or ring is a model of the appropriate algebraic theory. Algebraic theories are usually
\emph{presented} in terms of logical syntax, that is, as first-order theories whose signatures allow only functional
symbols, and whose axioms are universally quantified equations. In his seminal thesis,
\citet{lawvereFUNCTORIALSEMANTICSALGEBRAIC1963} defined a presentation-free categorical notion of universal algebraic
structure, called a Lawvere theory. Programming Languages, such as the $\lambda$-calculus, can be viewed as algebraic
structures with variable-binding operators, which can be formalised using second-order algebraic
theories~\cite{fioreSecondOrderAlgebraicTheories2010}, or algebraic theories with closed
structure~\cite{hylandClassicalLambdaCalculus2017}, called $\lambda$-theories, making the $\lambda$-calculus the
presentation of the initial $\lambda$-theory $\Lambda$. Our family of reversible languages $\PiLang$ have been presented
as first-order algebraic
2-theories~\cite{cohenCoherenceRewriting2theories2009,bekeCategorificationTermRewriting2011,yanofskySyntaxCoherence2000},
which are a categorification of algebraic theories. The types $\zerot$ and $\onet$ are nullary function symbols, the
type formers $+$ and $\times$ are binary function symbols, the 1-combinators are invertible reduction rules, and the
2-combinators are equations or coherence diagrams of compositions of reduction rules. Just like models of Lawvere
theories are given by algebras of (finitary) monads on $\SetCat$, models of 2-theories are given by algebras of 2-monads
on $\CatCat$. Our development is related to the free symmetric monoidal completion 2-monad.

@@

Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories
were studied by \citet{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a
groupoid and the proof of its universal property has been considered
by~\citet{piceghelloCoherenceMonoidalGroupoids2020}. Free commutative monoids in type theory have been studied by
\citet{gylterudMultisetsTypeTheory2020}, and using HoTT by \citet{choudhuryFinitemultisetConstructionHoTT2019}. The free
symmetric monoidal groupoid $\FSM[A]$ over a groupoid $A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be
presented as an algebraic 2-theory using 1-HITs. These HITs and the proof of their universal property have been
considered by \citet{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}. The proof of
the universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal
categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The
existence of the proof is folklore. We have produced a proof and formalised it in constructive type theory. A stronger
universal property that $\BFin$ is biinitial in $\term{RigCat}$, is in~\cite{elguetaGroupoidFiniteSets2021}.

@@

%%%%%%%%%%%%%%
\paragraph{Permutations} Finding formal systems for expressing various flavors of computable functions has been a major
focus of logic and computer science since its inception. Permutations, being at the core of reversible computing, are an
interesting class of functions, for which there are few formal systems. We develop such a system bringing in all the
associated benefits of syntactic calculi, notably, their calculational flavor. Instead of comparing two reversible
programs by extensional equality of the underlying bijective functions, a calculus offers more nuanced techniques that
can enforce additional intensional constraints on the desired equality relation.
